\section{Introduction}

Information security risk management is a central component of modern
organizational governance. Standards such as ISO/IEC~27005 and frameworks
such as NIST~SP~800-37 describe structured processes for identifying,
analysing, and treating risk, and regulatory and advisory bodies including
ENISA and the Norwegian National Security Authority (NSM) provide additional
guidance on how such processes should be implemented in practice. At the
same time, organisations have become increasingly dependent on digital
systems, and the growing complexity of those systems has made systematic
risk management a practical necessity rather than a purely administrative
exercise.

In this environment, the tools organisations use to support risk management
vary widely. They range from general-purpose spreadsheets used to maintain
ad-hoc risk registers, through single-purpose visual techniques such as
bowtie analysis, to full Governance, Risk, and Compliance (GRC) platforms
that integrate risk registers, control frameworks, compliance testing, and
reporting. The appropriateness of any given tool depends on the size and
maturity of the organisation, the frameworks it must align with, and the
resources available to configure and maintain the tool. Despite the
importance of this choice, there is limited comparative literature
evaluating how well widely available tools actually support the requirements
set out by established risk management frameworks.

\subsection{Project background}

This thesis was written as a Bachelor project at the Department of
Information Security and Communication Technology at NTNU, based on a
problem description supplied by the academic supervisor. The original
problem description covered both open-source and commercial risk management
tools. During the early phase of the project, the scope was narrowed to
\textit{open-source tools only}, a decision driven by practical
considerations of accessibility: open-source tools can be deployed,
configured, and tested in a controlled environment within the constraints
of a bachelor project, whereas commercial GRC platforms typically require
procurement, licensing agreements, or vendor-mediated demonstrations that
fall outside the resources available. The rationale for this scope change
is developed further in Section~\ref{sec:project-scope}.

\subsection{Project description}

This thesis evaluates a selection of open-source tools used to support
information security risk management. The evaluation combines a structured
criteria set, derived from established frameworks and from the guidance
provided in ENISA and NSM publications, with hands-on assessment of each
tool. Selected tools are subjected to in-depth walkthroughs against
realistic organisational scenarios representing small, medium, and large or
public-sector contexts. The resulting evidence is used to produce a
comparative analysis that identifies strengths, limitations, and the types
of organisational contexts in which each tool is most appropriate.

\subsection{Project scope}
\label{sec:project-scope}

The evaluation is scoped to open-source tools for information security risk
management. Commercial GRC platforms are out of scope, for the reasons
noted in the project background. The thesis does not aim to develop new
software, extend existing tools, or perform a comprehensive cost-benefit
analysis. Factors such as licensing model, ease of use, deployment effort,
and documentation maturity are considered as part of the evaluation, but
the primary focus is on functional capability, alignment with recognised
frameworks, and usability from the perspective of an organisation
implementing structured risk management.

The evaluation considers a selection of tools representative of the range
of open-source tool support available, from lightweight visual techniques
such as bowtie analysis through to full GRC platforms. The comparison is
framed against the guidance of ISO/IEC~27005, NIST~SP~800-37, OCTAVE
Allegro, FAIR, and the recommendations of ENISA and NSM, with particular
attention to how each tool supports or constrains the requirements
expressed by these frameworks.

\subsection{Research question}
\label{sec:research-question}

This thesis addresses the following primary research question:

\begin{quote}
\textit{To what extent do widely available open-source information security
risk management tools meet the requirements set out by established risk
management frameworks, and how does their suitability vary across different
organisational contexts?}
\end{quote}

This is supported by the following sub-questions:

\begin{enumerate}
    \item Which types of open-source risk management tools are available,
          and how can they be categorised?
    \item Which evaluation criteria, drawn from established frameworks and
          advisory guidance, are appropriate for comparing such tools?
    \item How well do the selected tools satisfy these criteria in
          practice?
    \item What strengths and limitations are observed when applying the
          tools to realistic organisational scenarios?
    \item Which tools are most appropriate for which organisational
          contexts (small and medium-sized enterprises, large enterprises,
          public sector)?
\end{enumerate}

\subsection{Project goals}

The project has four primary goals:
\begin{itemize}
    \item To establish a structured set of evaluation criteria grounded in
          ISO/IEC~27005, NIST~SP~800-37, OCTAVE Allegro, FAIR, and the
          guidance provided by ENISA and NSM.
    \item To identify and categorise a representative selection of
          open-source risk management tools.
    \item To evaluate each selected tool in depth against the criteria set,
          supported by hands-on deployment and testing.
    \item To produce a comparative analysis assessing the suitability of
          each tool across small and medium-sized enterprises, large
          enterprises, and public-sector organisations, and to offer
          recommendations for practice.
\end{itemize}

\subsection{Thesis structure}

The remainder of this thesis is structured as follows.
Chapter~\ref{chap:theoretical-background} establishes the theoretical
background, covering information security risk management and the
frameworks used as the basis for the evaluation criteria.
Chapter~\ref{chap:methodology} describes the research methodology,
including the tool selection strategy, the derivation of the evaluation
criteria, the evaluation approach, and the organisational scenarios used
to support the walkthroughs. Chapter~\ref{chap:tool-overview} provides an
overview of the tools considered and their categorisation.
Chapter~\ref{chap:practical-evaluation} presents the practical evaluation
of each selected tool. Chapter~\ref{chap:comparative} synthesises the
findings into a comparative analysis organised by evaluation criteria,
framework alignment, and organisational suitability. Chapter~\ref{chap:conclusion}
presents the conclusions, limitations, and suggestions for further work.